---
title: "ECON 121 FA24 Final Exam"
author: "Henri Runge"
output: pdf_document ## or html_document
---

```{r setup, include=FALSE}
# set the code chunk format
knitr::opts_chunk$set(echo = TRUE)
# clear the environment
rm(list=ls())
# load all packages from the course
library(tidyverse)
library(fixest)
library(marginaleffects)
library(dplyr)
load(url("https://github.com/HenriRunge/Coding-samples/raw/refs/heads/main/project_star.rds"))
load(url("https://github.com/HenriRunge/Coding-samples/raw/refs/heads/main/maimonides.rds"))
# you do not need any other packages for the exam.
# if you load other packages, we cannot help you 
# troubleshoot any resulting knitting problems.
```

## In-person codeword (disregard if taking the exam remotely)
metrics


## Question 1

```{r}
# Your code here

# Standardizing test scores
project_star <- 
  project_star %>%
  mutate(std_stest = scale(stest))

feols(std_stest ~ smale + swhite + sfree, data = project_star, vcov =~schoolid)
```

\newpage

## Question 2

The coefficients on the student's gender and whether they are eligible
for free lunch are statistically significant at the 5% level, whereas the one on the
students race only is at the 9% level, implying that it is not statistically 
significant at conventional levels. From the coefficients it follows, that
being male decreases a students standardized test score by a fifth of a standard 
deviation, being white increases it by 13.6% of a standard deviation and being 
eligible for free lunch decreases it by nearly half a standard deviation.

\newpage

## Question 3

```{r}
# Your code here

# Create a dummy for being below grade level
project_star <- project_star %>% mutate(
                          below_gl = ifelse(stest < "400", 1, 0))

# Running the logit regression
 logit_model <- feglm(below_gl ~ smale + swhite + sfree, data = project_star, vcov =~schoolid, 
        family = 'logit')
  
# Computing marginal effects
  avg_slopes(logit_model) 

# Computing odds ratios
  oddsratio <- exp(coef(logit_model))
  print(oddsratio)
```

\newpage

## Question 4

The largest risk factor for being below grade level is being eligibile for 
free lunch. This follows from the fact that it has the largest positive marginal
effect and the largest odds ratio above 1. The odds ratio tells us that being 
eligible for lunch increases the odds for being below grade level by 193%. The marginal
effect coefficient tells us that being eligible for lunch increases
the probability of being below grade level by 6.77 percentage points.

\newpage

## Question 5

```{r}
# Your code here

# Create dummy for small class 
project_star <- project_star %>% mutate(small_class = ifelse(cltype == "1", 1, 0))

feols(std_stest ~ small_class | schoolid, data = project_star, vcov =~classid)

# I was not sure if we should use cltype here or already create the dummy. Since
# we care about class size and not the TA, i decided to go with the dummy. Also 
# in question 10 it asks about the small class effect in question 5, which
# points toward the fact that we should use the dummy here already. I hope that makes
# sense and is correct.
```

\newpage

## Question 6

Being in the small class increases the standardized test score
by approximately one fifth of a standard deviation. The gap between free lunch
and non-free lunch students in question 1 was approximately half a standard deviation.
This implies that the gap here is approximately 40% of that at question 1.
(0.190294/0.476254 = 0.3995641) As we include school fixed effects and class type 
assignment at school level was random for both students and teachers, these effects 
can reasonably be seen as causal. 




\newpage

## Question 7

```{r}
# Your code here

# Create dummy for no experience
project_star <- project_star %>% mutate(tnoexper = ifelse(texp == "0", 1, 0))

# generate the interaction term
project_star <- project_star %>% mutate( tnoexperXsmall_class = tnoexper*small_class)

# Testing for heterogeneous effects across teachers with and without experience
model_1<-feols(std_stest ~ small_class + tnoexper + tnoexperXsmall_class | schoolid, data = project_star, vcov =~classid)
print(model_1)

hypotheses(model_1, "(small_class + tnoexperXsmall_class) / small_class = 1")
```

\newpage

## Question 8

Based on the p-value, which is larger than 0.05, we cannot 
reject the null hypothesis that small class has the same effect for teachers with and without 
experience at conventional levels. Based on our estimate, the effect of having 
a small class is 3.3 times larger for teachers without experience than those 
with experience. The regression results tell us that having a small class 
increases standardized test scores by 17% of a standard deviation for teachers
with experience and by approximately 73% of a standard deviation for teachers
without any experience. Moreover, having a teacher without experience decreases
standardized test scores by 12% of a standard deviation.

\newpage

## Question 9

```{r}
# Your code here
feols(std_stest ~ csize | schoolid, data = project_star, vcov =~classid)
```
\newpage

## Question 10

The results tell us that increasing class size by one student decreases
the standardized test score by 1.9% of a standard deviation. The most important
assumption to convert the "small class" effect in question 5 to a per student 
effect is that the marginal effect of increasing class size is linear
and constant, such that increasing class size by one student has the same effect 
for classes with 15 students and classes with 25 students.

\newpage

## Question 11

```{r}
# Your code here

feols(avg_verb ~ disadv, data = maimonides, vcov=~school_id)
```

\newpage

## Question 12

The slope coefficient tells us that increasing the share of students who 
are disadvantaged by one percentage points decreases the average verbal score
by 0.33 points. The coefficient would stay approximately the same, while the standard
error shrinks, if the individual-level model is homoskedastic in the errors
and beta is not heterogeneous across individuals.

\newpage

## Question 13

```{r}
# Your code here

# Create running variable
maimonides <- maimonides %>% mutate(running_x = grade_size-41)

# Restrict to observations within 25 students of the cutoff
maimonides_restriced <- maimonides %>%
  filter(running_x>= -25 & running_x <= 25)

# Discretize x into 5-student bins
maimonides_restriced <- maimonides_restriced %>%
  mutate(bin = floor(running_x / 5) * 5)

# Calculate bin-specific means for class_size and avg_verb
maimonides_mean <- maimonides_restriced %>%
  group_by(bin) %>%
  summarise(
    mean_class_size = mean(class_size, na.rm = TRUE),
    mean_avg_verb = mean(avg_verb, na.rm = TRUE)
  )

# Scatterplot for class_size
ggplot(maimonides_mean, aes(x = bin, y = mean_class_size)) +
  geom_point(color = "blue", size = 3) +
  labs(
    title = "Binned Scatterplot of Class Size",
    x = "Distance from Cutoff (x)",
    y = "Mean Class Size"
  ) +
  theme_minimal()

# Scatterplot for avg_verb
ggplot(maimonides_mean, aes(x = bin, y = mean_avg_verb)) +
  geom_point(color = "red", size = 3) +
  labs(
    title = "Binned Scatterplot of Average Verbal Test Scores",
    x = "Distance from Cutoff (x)",
    y = "Mean Average Verbal Score"
  ) +
  theme_minimal()

```

\newpage

## Question 14

As clearly shows in the binned scatterplot with class_size as the outcome, Maimonides'
rule does indeed influence class_size. At the cutoff at 40, there is a clear
discontinuity. In the language of instrumental variables, this relationship is
called the first stage relationship. The evidence is less clear when looking 
at the scatterplot with average test scores as the outcome. Being at the cutoff 
clearly does increase test scores compared to being slightly below, but the
test scores fall drastically when being shortly above the cutoff, which makes
the evidence less convincing, since class size shortly above the cutoff should still 
be smaller than slightly below. In the language of instrumental variables, this 
relationship is called the reduced form relationship. 
\newpage

## Question 15

```{r}
# Your code here

# first generate dummy for being above cutoff, plus
# interaction of dummy with distance
maimonides_restriced <- 
  maimonides_restriced %>%
  mutate(above = if_else(running_x>=0,1,0),
         aboveXrunning_x = above*running_x)

# first stage local linear regression with rectangular kernel and bw = 25 (doesnt
# need to be specified because I still use the restricted dataset)
feols(class_size ~ above + running_x + aboveXrunning_x, 
      data = maimonides_restriced, 
      vcov = ~~school_id)

# reduced form local linear regression with rectangular kernel and bw = 25
feols(avg_verb ~ above + running_x + aboveXrunning_x, 
      data = maimonides_restriced, 
      vcov = ~~school_id)

```

\newpage

## Question 16

The first stage result shows that being above the cutoff reduces average class_size
by 15 students. The reduced form results show that being above the cutoff increases
average verbal test scores by 2.3 points. To interpret these effects as causal 
effects of Maimonides' rule, we need to assume that the instrument of being above
the cutoff is independently assigned, that it is relevant and that the 
exclusion restriction holds. Random assignment can be tested by including pre-determined 
variables in the reduced form regression and checking if the coefficient changes, 
it should not. Relevance can be tested by looking at the statistical significance of the 
first stage regression. The exclusion restriction cannot be tested, but has to be thought about 
whether it is likely that they hold. In our case, it is reasonable to assume 
that Maimonides's rule only affects test scores through reducing class size and thus
we should be able to conclude that the exclusion restriction holds.
If we think that causal effects are heterogenous, which they are likely to be, 
since there are always-takers (e.g. school 51026), i.e. schools who reduced class 
size despite not having passed the cutoff, we also have to check the monotonicity 
assumption. Again this can only be thought about: It seems reasonable to assume 
that there are no defiers, i.e. schools who would decrease class size if they are 
below the cutoff but not decrease it if they are above.


\newpage

## Question 17

```{r}
# Your code here
# first stage local linear regression with rectangular kernel and bw = 25
feols(class_size ~ above + running_x + aboveXrunning_x, 
      data = maimonides_restriced, 
      vcov = ~~school_id)

# reduced form local linear regression with rectangular kernel and bw = 25, controlling
# for predetermined variables
feols(avg_verb ~ above + running_x + aboveXrunning_x + disadv, 
      data = maimonides_restriced, 
      vcov = ~~school_id)

```

\newpage

## Question 18

2.294985/-15.053213
= -0.1524581

This result represents the average causal effect among grades that
were induced to reduce their class size due to Maimonides' rule, i.e. compliers. 
Since we already made the monotonicity assumption in Question 16, no further
assumptions are needed for this interpretation. It does not reflect the effects of
passing the cutoff among those schools that reduced class size anyway (always takers),
nor the effects of passing the cutoff among schools that never
would reduce their class size(never takers).

\newpage

## Question 19

```{r}
# Your code here
feols(avg_verb ~ running_x + aboveXrunning_x | class_size ~ above, 
      data = maimonides_restriced, 
      vcov = ~school_id)
feols(avg_verb ~ class_size, data = maimonides, vcov = ~school_id)
```

\newpage

## Question 20

As follows from the regression results and the t-statistic of -1.57, the TSLS 
result is not significantly different from zero. Compared to the OLS
regression, the coefficient changes sign from positive to negative and increases
approximately 5 times in magnitude. The difference could be due to endogeneity bias
in class size for the OLS regression: Schools in poorer neighborhoods, for example, 
might have larger class sizes and lower outcomes due to resource constraints, 
creating a positive bias in OLS. The TSLS uses an exogenous instrument
(Maimonides' rule) to remove this endogeneity bias and thus the
coefficient changes.






