---
title: "ECON 121 FA23 Problem Set 2"
author: "Henri Runge"
output: pdf_document # change to html_document if you don't have LaTeX
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # sets the code chunk format
rm(list = ls()) # clears the environment
```

## Question 1

Verbal: list group members.
Timo van Heek
\newpage

## Question 2

Verbal answer only, no code.
The Mincerian Wage Equation establishes a relationship between the dependent 
variable, the logarithm of the hourly wage, and the explanatory variables educ-
ation and experience as well as experience squared. Given the log-level 
relationship and assuming that experience and education are exogenous, one can 
say that a one-year increase in education increases the hourly wage by 
100*beta1%, holding the years of experience fixed. Experience is featured twice 
also including a squared term, because intuitively we might expect that the 
effect of experience is different for the first years of experience compared to 
the years of experience in later years of one's career. Depending on the sign of 
the squared term, it could either be that the return to experience increases 
with years of experience (positive sign on squared term) or decreases (negative 
sign on squared term). Labor economics typically assumes that the sign is 
positive so that the wage of young professionals is lower and increases more
steeply in later years; this is referred to as "deferred compensation" and 
supposed to incentivize workers to stay in the company, work hard and not shirk.

\newpage

## Question 3

Code: Load packages and dataset, modify sample, generate variables, summarize data.

Verbal: Interpret the summary statistics.

```{r message=FALSE, warning=FALSE}
# The PDF will show the code you write here but not the output.
# Load packages and dataset, modify sample, generate variables here.
# Load the CPS data
load (url("https://github.com/HenriRunge/Coding-samples/raw/refs/heads/main/cps18.rds"))
# Drop individuals with fewer than 50 weeks worked, fewer than 35 hours/week, or 0 earnings
fil_data <- cps18 %>%
filter(wkswork >= 50, hrs_per_wk >= 35, incwage > 0) %>%
  
# Generate log hourly wage variable
mutate(hourly_wage = incwage / (wkswork * hrs_per_wk),
         log_hourly_wage = log(hourly_wage)) %>%  
  
# Create race/ethnicity dummies
mutate(Black = ifelse(race == "black", 1, 0),
         Native = ifelse(race == "native", 1, 0),
         Asian = ifelse(race == "asian/pacific", 1, 0),
         Other_Multiple = ifelse(race == "multiple/other", 1, 0)) %>%
  
# Create education variable (adjust categories to years of schooling)
mutate(education_years = case_when(
    ed_lt_hs == 1 ~ 8,            
    ed_some_hs == 1 ~ 10,
    ed_hs_degree == 1 ~ 12, 
    ed_some_col == 1 ~ 14,     
    ed_ba_degree == 1 ~ 16,          
    ed_post_degree == 1 ~ 18,          
    TRUE ~ NA_real_)) %>%
  
# Generate potential experience and squared experience
mutate(experience = age - education_years - 5,
         experience_sq = experience^2) 

```

```{r}
# The PDF will show the code AND output here.
# Summarize the data here.
  
summary(fil_data)

# Something interesting to observe is that the mean of ed_some_col is the 
# highest among the education dummy variables, implying that most people in the 
# dataset obtained some college education. Moreover, the annual wage differs 
# between $4 and above $1.6 million, where it seems that something most have gone 
# wrong with the former observation.
 fil_data %>%
  arrange(incwage) %>%    # Sort in ascending order by incwage
  slice(1:10)
# It seems there are some people who worked quite a lot but did basically earn 
# no money at all. It might be that they just did voluntary work, but to me it 
# seems unlikely that someone would work 70 hours a week for 52 weeks (like the 
# sixth observation in the table below) without earning any money and thus my 
# interpretation would be that they misinterpreted the question on incwage and 
# filled in their hourly wage. 
```

\newpage

## Question 4

Code: Estimate regression.

Verbal: Interpret your results.

```{r}
# All question 4 code here
# Estimate the Mincerian Wage Equation 
mincer_fixest <- feols(log_hourly_wage ~ education_years + experience + 
                         experience_sq, data = fil_data, vcov = 'hetero')

# Display the regression output
summary(mincer_fixest)

# The estimations imply that, given years of experience, one additional year of 
# education increases the hourly wage by approximately 13.8%. Moreover, given 
# years of education, one additional year of experience changes the hourly wage 
# according to this equation: 0.03-0.0004experience. Thus, the first year of 
# experience increases the hourly wage by approximately 3% and, for instance, the 
# 20th year increases it by (100*(0.03-0.0004*20))%=2.2%. Lastly, without any 
# education and experience the expected wage would be $2.05, which clearly is an 
# extrapolation and should not be interpreted as carrying any real-world meaning.
```

\newpage

## Question 5

Code: Estimate regression.

Verbal: Interpret your results.

```{r}
# All question 5 code here
# Estimate the extended Mincerian Wage Equation with race and sex controls 
mincer_extended <- feols(log_hourly_wage ~ education_years + experience + 
                           experience_sq + Black + Native + Asian + 
                           Other_Multiple + male, 
                           data = fil_data, vcov = 'hetero')

# Display the regression output
summary(mincer_extended)

# The return to education slightly increases when controlling for race and sex. 
# This suggests omitted variable bias, where the education variable captures
# some of the effect of sex and ethnicity on the hourly wage. 
```

\newpage

## Question 6

Code: Assess statistical significance.

Verbal: Interpret your results.

```{r}
# All question 6 code here
hypotheses(mincer_extended, "Black = -male")

# Given our high z-statistic and the correspondingly low p-value, we can 
# conclude that the black-white log wage gap is statistically different from
# the female-male log wage gap.
```


\newpage

## Question 7

Code: Estimate regressions and assess significance of differences between them.

Verbal: Interpret your results.

```{r}
# All question 7 code here
# Subset the data for men (male == 1) and women (male == 0)
data_men <- fil_data %>% filter(male == 1)
data_women <- fil_data %>% filter(male == 0)

# Extended regression model for men
model_men <- feols(log_hourly_wage ~ education_years + experience + experience_sq + 
                   Black + Native + Asian + Other_Multiple, data = data_men, vcov = 'hetero')

# Extended regression model for women
model_women <- feols(log_hourly_wage ~ education_years + experience + experience_sq + 
                     Black + Native + Asian + Other_Multiple, data = data_women, vcov = 'hetero')

# From the regression results it follows that the return to education is 
# about 1 percentage point higher for women than for men. (14.8% vs. 13.9% 
# increase in the hourly wage for one additional year of schooling, keeping 
# everything else constant).

# Extract coefficients for education_years
coef_men <- coef(model_men)["education_years"]
coef_women <- coef(model_women)["education_years"]

# Extract standard errors directly from the summary
se_men <- summary(model_men)$coeftable["education_years", "Std. Error"]
se_women <- summary(model_women)$coeftable["education_years", "Std. Error"]

# Print coefficients and standard errors
cat("Return to education for men:", coef_men, "\n")
cat("Standard Error for men:", se_men, "\n\n")

cat("Return to education for women:", coef_women, "\n")
cat("Standard Error for women:", se_women, "\n")
# Compute the Z-statistic for the difference
z_stat_gender <- (coef_women - coef_men) / sqrt(se_men^2 + se_women^2)

# Compute the p-value
p_value_gender <- 2 * (1 - pnorm(abs(z_stat_gender)))

# Print the Z-statistic and p-value
cat("Z-statistic:", z_stat_gender, "\n")
cat("P-value:", p_value_gender, "\n")

# The findings suggest that the difference in return to education for
# men and women is statistically different, as the null hypothesis that 
# education_years is equal for men and women is rejected at a 
# significance level of alpha = 0.05.
```

\newpage

## Question 8

Code: Estimate regression, use delta method.

Verbal: Interpret your results.

```{r}
# All question 8 code here
# Extended regression with interaction terms for male and education
model_interaction <- feols(log_hourly_wage ~ education_years + experience + experience_sq + 
                     Black + Native + Asian + Other_Multiple + male + education_years*male + experience*male + experience_sq*male + Black*male + Native*male + Asian*male + Other_Multiple*male, data = fil_data, vcov = 'hetero')

# Displaying the results
etable(model_interaction)

# The results are approximately the same with the coefficient on education_years 
# which gives the slope for women being 0.1485 and the slope for men being
# 0.1485-0.0097=0.1388.

# Estimating the ratio of returns for women to men using the delta method
hypotheses(model_interaction, "education_years / (education_years + education_years:male) = 1")


```

\newpage

## Question 9

Code: Load dataset, compute unweighted and weighted means.

Verbal: Interpret your results.

```{r}
# All question 9 code here

```

\newpage

## Question 10

Code: Modify sample, estimate unweighted and weighted regressions.

Verbal: Assess unweighted versus weighted results.

```{r}
# All question 10 code here

```

\newpage

## Question 11

Verbal answer only, no code.

\newpage

## Question 12

Verbal answer only, no code.

\newpage

## Question 13

Code: Estimate regression.

Verbal: Explain choice of covariates, interpret results.

```{r}
# All question 13 code here

```

