---
title: "ECON 121 FA23 Problem Set 3"
author: "Henri Runge"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # sets the code chunk format
rm(list = ls()) # clears the environment
```

## Question 1

Verbal: list group members.

Timo van Heek


Dánjal Augustinussen

\newpage

## Question 2

Code: Load packages and dataset,generate variables, summarize data.

Verbal: Interpret the summary statistics.

```{r message=FALSE, warning=FALSE}
# The PDF will show the code you write here but not the output.
# Load packages and dataset, generate variables here.

load (url("https://github.com/HenriRunge/Coding-samples/raw/refs/heads/main/nhis2010.rds"))
library(ggplot2)
library(dplyr)
library(tidyverse)
library(fixest)
library(marginaleffects)
options(scipen=999)
nhis2010 <- nhis2010 %>% 
mutate(low_health = ifelse(health %in% c("Fair", "Poor"), 1, 0))

```

```{r echo=TRUE}
# The PDF will show the code AND output here.
# Summarize the data here.
summary(nhis2010)

# The results show that 16.14% of the population reported having fair or poor
# health in the sample, implying that most individuals reported having good, very
# good or excellent health, the average age was 50.79 years with individuals 
# from ages 25 to 85 being included in the sample, 43.81% of the sample were male,
# and 57.64% were white, 16.11% black, 18.24% hispanic, 6.249% asian and 1.757% from
# another ethnicity. The majority of the sample was married with the next largest
# group being those that have never been married and the average education was 13.81
# years with a maximum of 19 years and a low of 1 year. Moreover, most people
# in the sample were employed with the second largest group being not in the labor 
# force, the family income is similar across bins except the largest bin being 
# the lowest one with incomes up to $34,999. The average BMi was 26.69, 12.89% of
# the sample had died by 2019, 17.44% were uninsured, the average hours of sleep 
# reported were 7.158, most people reported never having felt sad in the past 30 days,
# 9.488% had cancer before, 3.798% had a heart attack before, 35.71% had hypertension
# before, 12.72% have had diabetes, the mean days with 5 or more drinks in the
# past year were 11, 42.02% had smoked over 100 cigarettes in their life and the
# average amount of 10+ minutes vigorous activity per week was 1.494.
```

\newpage

## Question 3

Code: Draw graph with two line plots.

Verbal: Interpret.

```{r echo=TRUE}
# All question 3 code here.
# Subset the data for people with poor/ fair health (low_health == 1) and good 
# health (low_health == 0)
data_low_health <- nhis2010 %>% filter(low_health == 1)
data_high_health <- nhis2010 %>% filter(low_health == 0)

# Calculate mortality rate by age for fair-to-poor health group
mortality_by_age_low_health <- data_low_health %>%
  group_by(age) %>%
  summarize(mortality_rate = mean(mort, na.rm = TRUE))

# Calculate mortality rate by age for good-to-excellent health group
mortality_by_age_high_health <- data_high_health %>%
  group_by(age) %>%
  summarize(mortality_rate = mean(mort, na.rm = TRUE))

# Plot mortality rates by age for the two groups
ggplot() +
  geom_line(data = mortality_by_age_low_health, aes(x = age, y = mortality_rate, color = "Low Health")) +
  geom_line(data = mortality_by_age_high_health, aes(x = age, y = mortality_rate, color = "High Health")) +
  labs(title = "Mortality Rate by Age for Different Health Status Groups",
       x = "Age", y = "Mortality Rate",
       color = "Health Status") +
  theme_minimal()
```

The risk of death increases with age as one would expect. From age 40 onward the
low-health line lies consistently above the high-health line implying greater
risk of death for those who self-reported having low health. Before age 40 there
are some years in which the high health line is above the low health line but 
generally the same picture obtains. The "outlier" or inconsistent years probably
occur because you usually do not die because of natural health related reasons in 
such young years, which is why other reasons for death become more important and 
a car accident for instance can hit either low or high health people.
\newpage

## Question 4

Code: Draw bar graphs.

Verbal: Interpret your results.

```{r echo=TRUE}
# All question 4 code here

# Calculate mortality rate by family income level
mort_by_incfam <- nhis2010 %>%
  drop_na(incfam, mort) %>%
  group_by(incfam) %>%
  summarize(mortality_rate = mean(mort, na.rm = TRUE))

# Calculate fair/poor health rate by family income level
low_health_by_incfam <- nhis2010 %>%
   drop_na(incfam, health) %>%
  group_by(incfam) %>%
  summarize(fair_poor_health_rate = mean(low_health, na.rm = TRUE))

# Plot mortality by family income level
ggplot(mort_by_incfam, aes(x = incfam, y = mortality_rate)) +
  geom_bar(stat = "identity") +
  labs(title = "Mortality Rate by Family Income Level",
       x = "Family Income Level",
       y = "Mortality Rate") +
  theme_minimal()

# Plot fair/poor health by family income level
ggplot(low_health_by_incfam, aes(x = incfam, y = fair_poor_health_rate)) +
  geom_bar(stat = "identity") +
  labs(title = "Fair/Poor Health Rate by Family Income Level",
       x = "Family Income Level",
       y = "Fair/Poor Health Rate") +
  theme_minimal()

# b) Create education categories 
nhis2010 <- nhis2010 %>%
  mutate(education_level = case_when(
    edyrs < 12 ~ "1. less than high school",
    edyrs == 12 ~ "2. high school",
    edyrs >= 13 & edyrs <= 15 ~ "3. some college",
    edyrs == 16 ~ "4. college degree",
    edyrs > 16 ~ "5. post-graduate"
  ))

# Mortality by education level
mortality_education <- nhis2010 %>%
  drop_na(education_level, mort) %>%
  group_by(education_level) %>%
  summarize(mortality_rate = mean(mort, na.rm = TRUE))

# (d) Fair/poor health by education level
health_education <- nhis2010 %>%
  drop_na(education_level, health) %>%
  group_by(education_level) %>%
  summarize(low_health_rate = mean(low_health, na.rm = TRUE))

# Plot mortality by education level
ggplot(mortality_education, aes(x = education_level, y = mortality_rate)) +
  geom_bar(stat = "identity") +
  labs(title = "Mortality Rate by Education Level", x = "Education Level", y = "Mortality Rate") +
  theme_minimal()

# Plot low health by education level
ggplot(health_education, aes(x = education_level, y = low_health_rate)) +
  geom_bar(stat = "identity") +
  labs(title = "Fair/Poor Health Rate by Education Level", x = "Education Level", y = "Fair/Poor Health Rate") +
  theme_minimal()

```
The first graph showing the mortality rate by family income level
shows that the higher the family income, the lower the mortality rate. 
The highest mortality rate is reported for those with a family income of
between $0 and $34,999 with approximately 18.75%. Thereafter, the mortality
rate gradually declines down to its low at about 5% for those with a family
income of $100,000 and over.
The second graph shows the fair/poor health rate by family income level 
and gives a similar picture to the first graph with the fair/poor health rate
being higher for those with low family income level. The highest fair/poor 
health rate is reported for those with a family income of between $0 and $34,999 
with approximately 27%. Thereafter, the mortality rate gradually declines down 
to its low at about 4% for those with a family income of $100,000 and over.
The third graph shows the mortality rate by education level and again draws 
a picture where higher education generally is associated with a lower mortality
rate.The maximum mortality rate is obtained for individuals with years of 
schooling corresponding to less than a high school degree at approximately 21%, 
whereas the low obtains for those with years of schooling corresponding to a
college degree at slightly below 7.5%. Interestingly, the mortality rate slightly
increases fo individuals with a post-graduate level of schooling to slightly above 
7.5%.
The fourth graph displaying the fair/poor health rate by education level shows that
the incidence of self-reported fair/poor health decreases with education level.
The high is obtained at approximately 33% for individuals with years of education
corresponding with less than a high school diploma. Thereafter, the fair/poor
health rate steeply decreases to its low for individuals with post-graduate 
education at about 5%.

\newpage

## Question 5

Code: Estimate regressions.

Verbal: Interpret and compare.

```{r echo=TRUE}
# All question 5 code here

# Create age string variable
nhis2010 <- nhis2010 %>%
  mutate(agegroup = case_when((25<= age & age <= 29) ~ "25-29",
                             (30<= age & age <= 34) ~ "30-34",
                             (35<= age & age <= 39) ~ "35-39",
                             (40<= age & age <= 44) ~ "40-44",
                             (45<= age & age <= 49) ~ "45-49",
                             (50<= age & age <= 54) ~ "50-54",
                             (55<= age & age <= 59) ~ "55-59",
                             (60<= age & age <= 64) ~ "60-64",
                             (65<= age & age <= 69) ~ "65-69",
                             (70<= age & age <= 74) ~ "70-74",
                             (75<= age & age <= 79) ~ "75-79",
                             (80<= age & age <= 85) ~ "80-85"))
                             
# Linear Probability Model  
lpm_mortality <- feols(mort ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero')

lpm_health <- feols(low_health ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero')

# Probit Model 
probit_mortality <- feglm(mort ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero', family = 'probit')

probit_health <- feglm(low_health ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero', family = 'probit')

# Logit Model 
logit_mortality <- feglm(mort ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero', family = 'logit')

logit_health <- feglm(low_health ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero', family = 'logit')

# Summary of the regression results
summary(lpm_mortality)
summary(lpm_health)
summary(probit_mortality)
summary(probit_health)
summary(logit_mortality)
summary(logit_health)

# Marginal effects
avg_slopes(probit_mortality)
avg_slopes(probit_health)
avg_slopes(logit_mortality)
avg_slopes(logit_health)
           
# Odds ratio
mortality_odds_ratios <- exp(coef(logit_mortality))
health_odds_ratios <- exp(coef(logit_health))
print(mortality_odds_ratios)
print(health_odds_ratios)

# Create a table to be able to compare LPM, probit and logit results
coef_lpm_mort <- coef(lpm_mortality)
coef_probit_mort <- coef(probit_mortality)
coef_logit_mort <- coef(logit_mortality)
cbind(coef_lpm_mort, coef_probit_mort, coef_logit_mort)


coef_lpm_health <- coef(lpm_health)
coef_probit_health <- coef(probit_health)
coef_logit_health <- coef(lpm_health)
cbind(coef_lpm_health, coef_probit_health, coef_logit_health)
```

I chose to include the variables age and education both in the form of string variables 
because both age (I chose 5 years as the bin width to allow for enough detail)
and education are likely to not have linear effects on mortality 
and health. For age, the graph drawn for question 3, suggests a quadratic relationship, 
but because we only have individuals age 25-85 the interpretation of the coefficients
would be difficult. In terms of results the table comparing the coefficients
of the LPM, the probit and the logit model shows that the coefficients for the 
mortality and health regression differ in terms of magnitude between the models, 
but mostly have the same sign; another observation is that for many coefficients 
the one of the logit model is greater in terms of magnitude than the one of the 
probit model and the one of the probit model larger than the one of the linear 
probability model. In terms of marginal effects the ones of the logit 
model are approximately twice the size of those of the probit model. Lastly, 
we have the odds ratios which for instance show that, compared to 25-29 year olds, 
70-74 year olds have a 27 times higher odd of dying, etc.



\newpage

## Question 6

Code: Use the results from the mortality logit model to compare the two groups.

Verbal: Interpret your results.

```{r echo=TRUE}
# All question 6 code here


# Create dummy variables for incfam categories
nhis2010 <- nhis2010 %>%
  mutate(
    incfam_0_34k = ifelse(incfam == "$0 - $34,999", 1, 0),
    incfam_35_49k = ifelse(incfam == "$35,000 - $49,999", 1, 0),
    incfam_50_74k = ifelse(incfam == "$50,000 - $74,999", 1, 0),
    incfam_75_99k = ifelse(incfam == "$75,000 - $99,999", 1, 0),
    incfam_100k_plus = ifelse(incfam == "$100,000 and over", 1, 0)
  )

# Create dummy variables for education levels
nhis2010 <- nhis2010 %>%
  mutate(
    less_than_high_school = ifelse(edyrs < 12, 1, 0),
    high_school = ifelse(edyrs == 12, 1, 0),
    some_college = ifelse(edyrs >= 13 & edyrs <= 15, 1, 0),
    college_degree = ifelse(edyrs == 16, 1, 0),
    post_graduate = ifelse(edyrs > 16, 1, 0)
  )

# Fit the Logit model with dummy variables for income and education
logit_mortality_new <- feglm( mort ~ agegroup + high_school + some_college + college_degree + post_graduate +
    incfam_35_49k + incfam_50_74k + incfam_75_99k + incfam_100k_plus + black + hisp 
                       + asian + other, data = nhis2010, vcov = 'hetero', family = 'logit')


# Extract coefficients from the logit model
coeffs <- coef(logit_mortality_new)

# Calculate the log-odds for Group A (Asian, less than high school, income <$35,000)
log_odds_A <- coeffs["(Intercept)"] + 
              coeffs["asian"] 
            

# Convert log-odds to probability for Group A
prob_A <- exp(log_odds_A) / (1 + exp(log_odds_A))

# Calculate the log-odds for Group B (Black, college degree, income >$100,000)
log_odds_B <- coeffs["(Intercept)"] + 
              coeffs["black"] + 
              coeffs["college_degree"] + 
              coeffs["incfam_100k_plus"] 
              
# Convert log-odds to probability for Group B
prob_B <- exp(log_odds_B) / (1 + exp(log_odds_B))


hypotheses(logit_mortality_new, "(asian)-(black+college_degree+incfam_100k_plus)=0")
```
The results show that at the same age Asian adults with less than 12 years of 
education and family incomes less than $35k have a greater mortality risk than
Black adults with 16 years of education and family incomes over $100k. This difference
is statistically significant. And no, this model is not the best for testing 
the differences between these groups. We should include interaction terms between the 
ethnicity variables and income as well as education levels to see if the effect
of education levels changes with ethnicity and the same for income.


\newpage

## Question 7

Verbal: Assess causality.
The coefficients on family income should not be thought of as causal, because 
observational data like the data from the National Health Interview Survey might
be subject to confounding factors and/ or endogeneity issues. Despite controlling
for some other socioeconomic variables in our regression, the coefficients on 
family income might capture the effect of other variables influencing both
family income and health, such as ability, possibly leading to higher income
and better health because of better life choices or family background, where 
children of doctors might have better health and better family income later in their 
lives. Moreover, health issues could impact one's ability to work and thus diminish
family earnings, creating reverse causaliy. To establish causality, we would need 
to use methods designed to infer causality, such as diff-in-diff approaches based
on natural experiments.
\newpage

## Question 8

Code: Assess how much health behavior can explain the mortality logit results.

Verbal: Interpret your results.

```{r echo=TRUE}
# All question 8 code here

# Extended Logit Model with health insurance, and health behaviors
logit_extended <- feglm(mort ~ agegroup + education_level + incfam + black + hisp 
                       + asian + other + uninsured + alc5upyr + smokev + vig10fwk + hrsleep , 
                       data = nhis2010, vcov = 'hetero', family = 'logit')

# Calculate Odds Ratio for extended model
extended_odds_ratios <- exp(coef(logit_extended))

# Display odds ratio
print(extended_odds_ratios)
print(mortality_odds_ratios)
```
The education level odds ratios all move closer to 1 in the extended model, implying
that the mediating variables - whether one is uninsured, the days one had more than 5 drinks,
whether one ever smoked more than 100 cigarettes, the frequency of vigorous 
activity per week and the usual hours of sleep per night - explain quite a lot
of the relationship between education and mortality. This is because an odds ratio
close to 1 implies that the coefficient on a particular variables is 0, implying that
it has no relation to the independent variable and thus only captured the effect
of the mediating variables in the short regression.For the family income level, 
however, three of four levels actually move further from 1. This implies that here 
the mediating variables cannot explain the relationship between family income
and mortality.