---
title: "ECON 121 FA23 Problem Set 4"
author: "Henri Runge"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # sets the code chunk format
rm(list = ls()) # clears the environment
```

## Question 1

Verbal: list group members.

Danjál Augustinussen



Timo van Heek


\newpage

## Question 2

Code: Load packages and dataset, summarize data.

Verbal: Interpret the summary statistics.

```{r message=FALSE, warning=FALSE}
# The PDF will show the code you write here but not the output.
# Load packages and dataset here.
library(fixest)
library(tidyverse)
library(dplyr)
library(marginaleffects)
load (url("https://github.com/HenriRunge/Coding-samples/raw/refs/heads/main/nlsy_kids.rds"))

```

```{r}
# The PDF will show the code AND output here.
# Summarize the data here.
summary(nlsy_kids)

# Summarize pre-Head Start characteristics by Head Start participation
summary_table <- nlsy_kids %>%
  group_by(head_start) %>%
  summarize(
    mean_mom_edu = mean(momed, na.rm = TRUE),
    mean_income = mean(lninc_0to3, na.rm = TRUE),
    mean_lnbw = mean(lnbw, na.rm = TRUE),
    prop_black = mean(black, na.rm = TRUE),
    prop_hispanic = mean(hispanic, na.rm = TRUE),
    prop_male = mean(male, na.rm = TRUE),
    mean_dad_home = mean(dadhome_0to3, na.rm = TRUE),
    mean_ppvt = mean(ppvt_3, na.rm = TRUE),
    mean_firstborn = mean(firstborn, na.rm = TRUE),
    
    n = n()
  )

transposed_table <- as.data.frame(t(summary_table))
print(transposed_table)


# Calculating summary statistics, it shows that of the 881 children who participate in Head
# Start have mothers with on average a quarter of a year less of education attainment.
# Moreover, the logarithm of family income between ages 0 and 3 is lower for participants.
# Birth weight is approximately the same for both groups. Of the participants a much
# larger portion of close to 52% is black compared to only 27% for non-participants.
# Conversely, the portion of non-participants being hispanic is slightly larger at 20% 
# compared to 19%. Participants are 53% male, whereas non-participants are 50% male.
# Dad's of non-participants were at home more of the time than those of participants.
# Basically, for participants on average fathers were at home 50% of the time between
# ages 0 and 3, whereas for non-participants they were home 72% of the time.
# Second to last, the mean PPVT score was 4 points higher for non-participants.
# Lastly, the percentage of participants being first-born children was slightly higher
# at 42% compared to 40%.
```

\newpage

## Question 3

Code: Regression.

Verbal: Interpret.

```{r}
# All question 3 code here.
feols(comp_score_5to6 ~ head_start, data = nlsy_kids, vcov = ~mom_id)

# Summarize the variables of interest
summary(nlsy_kids$head_start)
summary(nlsy_kids$comp_score_5to6)

# Calculate standard deviation of test scores
sd(nlsy_kids$comp_score_5to6, na.rm = TRUE)

# If Head Start participation is considered to be exogenous, we can conclude
# that it reduces composite test scores by 5.8 points. Given the standard deviation 
# of test scores, this implies a reduction of approximately a quarter of a standard 
# deviation. However, it is not reasonable to assume that Head Start participation
# is exogenous, because it is likely that there are omitted variables at the 
# family level, such as household income, racial ethnicity, the mother's 
# years of education etc. Omitting these variables will most likely bias the estimated 
# coefficient downward. I am assuming that it would turn positive when controlling for 
# these variables, because Head Start participants are disadvanteged. Thus, right now 
# the coefficient implies a negative effect of participation on test scores because 
# it captures the negative effects of low income, lower education of mothers, etc. 
# on test scores. Given that a child is disadvantaged 
# (controlling for the mentioned variabels), the participation in Head Start will most 
# likely show a positive effect on test scores.

```

\newpage

## Question 4

Code: Regression.

Verbal: Interpret.

```{r}
# All question 4 code here

# Group data by mother ID and calculate family means
family_means <- nlsy_kids %>%
  group_by(mom_id) %>%
  summarize(mean_head_start = mean(head_start, na.rm = TRUE),
    mean_test_score = mean(comp_score_5to6, na.rm = TRUE))
# Fit an OLS regression using the family means
feols(mean_test_score ~ mean_head_start, data = family_means, vcov = 'hetero')

# The estimated coefficient is a bit less than twice as large as the one 
# computed at question 3. Looking at "between mother" variation, the effect
# of participation in Head Start on test scores appears to be slightly less than 
# one half of a standard deviation. The problem with this regression it that it 
# removes within-family variation, making the coefficient more sensitive to biases
# related to omitted variables at the family level and therefore biasing it 
# even further downward.
```

\newpage

## Question 5

Code: Regression.

Verbal: Interpret.

```{r}
# All question 5 code here

# Estimate the fixed-effects model
feols(comp_score_5to6 ~ head_start | mom_id, data = nlsy_kids)

# The results of the fixed-effects model imply that Head Start participation
# increases test scores by approximately one third of a standard deviation when controlling
# for between-family variation. The explanation to the change in sign has basically 
# already been given at the previous questions. The omitted variables related
# to between-family variation biased the coefficient downward. Now that they are
# being controlled for, the coefficient reflects the actual effect of Head Start
# participation on test scores. This is the case because before it also captured
# the fact that Head Start participants predominantly come from disadvantaged families,
# which (logically) seems to decrease test scores. This, however, has nothing
# to do with the effect of Head Start on test scores, it is a confounder. Thus,
# the estimate that results from the fixed-effects model most likely reflects
# the effect of Head Start on test scores.
```

\newpage

## Question 6

Code: Regression.

Verbal: Interpret.

```{r}
# All question 6 code here

# Creating a subset with no missing values for the multiple variables
q6_subset <- nlsy_kids %>%
  drop_na(head_start, male, firstborn, lninc_0to3, dadhome_0to3, lnbw)

# Running FE regression with and without pre-Head Start variables
# Model without pre-Head Start variables
feols(comp_score_5to6 ~ head_start | mom_id, data = q6_subset)

# Model with pre-Head Start variables
feols(comp_score_5to6 ~ head_start + male + firstborn + 
                      lninc_0to3 + dadhome_0to3 + lnbw | mom_id, data = q6_subset)

# To be able to compare the multipe with the simple regression model, I firstly
# create a dataset without any missing values for the covariates that I want to include.
# I chose to include the variables on whether an individual is the first born child, 
# its household's income ages 0 to 3, whether the individual is male and the birthweight 
# as pre-Head Start covariates. I wanted to include the early childhood test score 
# as well as it is determined prior to Head Start participation and captures
# within-family variation. However, including it changes the coefficient on Head 
# Start dramatically. Checking this odd result, it shows that many individuals 
# have na scores for the early childhood variable, apparently changing the composition
# of the subset of the dataset being used for this regression dramatically. 
# Thus, the dataset for this regression is not at all comparable to the one used for question 5. 
# To be able to compare the results better, I excluded this variable.  
# Generally, for variables to qualify as pre-Head Start covariates in a mother-fixed-effect
# model they must have been determined pre-Head Start and they cannot reflect between-family 
# variation. Thus, momeduc and the racial ethnicity dummies are disqualified because
# they reflect between-family variation, which is captured by including fixed effects.
# Moreover, the variables comp_score_7to10 to fphealth are disqualified because they
# are only determined after Head Start participation. The variables included in my model, however, 
# all reflect within-family variation not captured by including mother-fixed effects and 
# were determined prior to Head Start participation. When including pre-Head Start 
# covariates in the fixed-effects model, the estimated coefficient on the Head Start
# variable is reduced slightly in magnitude but keeps the same sign. In the multiple regression,
# Head Start participation appears to increase test scores by approximately one quarter of a standard
# deviation, whereas it's slightly more than that in the simple regression. Thus, we can 
# conclude  that the fixed effect estimate of the effect of Head Start is robust to include 
# pre-Head Start covariates.
```

\newpage

## Question 7

Code: Regressions.

Verbal: Interpret.

```{r}
# All question 7 code here

# Standardize test scores for all age groups
score_subset <- nlsy_kids %>%
  mutate(
    comp_score_5to6_std = scale(comp_score_5to6, center = TRUE, scale = TRUE),
    comp_score_7to10_std = scale(comp_score_7to10, center = TRUE, scale = TRUE),
    comp_score_11to14_std = scale(comp_score_11to14, center = TRUE, scale = TRUE)
  ) %>%
  drop_na(head_start, comp_score_5to6, comp_score_11to14, comp_score_7to10)
  

# Fixed-effects model for test scores at ages 5-6
feols(
  comp_score_5to6_std ~ head_start | mom_id,
  data = score_subset
)
# Fixed-effects model for test scores at ages 7-10
feols(
  comp_score_7to10_std ~ head_start | mom_id,
  data = score_subset
)

# Fixed-effects model for test scores at ages 11-14
feols(
  comp_score_11to14_std ~ head_start | mom_id,
  data = score_subset
)

# As follows from the regression results, the effects of Head Start on test scores
# fade out with age. In both regressions the estimated coefficient is closer
# to zero. However, the results indicate that the effects don't monotonically decrease
# with age as the effect for 11 to 14 year olds is larger than the one for 7 to 10
# 10 year olds. Generally, the effect remains significantly positive in later years of 
# childhood.
```

\newpage

## Question 8

Code: Regressions.

Verbal: Interpret.

```{r}
# All question 8 code here
# Fixed-effects model for repeating a grade
feols(
  repgrade ~ head_start | mom_id,
  data = nlsy_kids
)

# Fixed-effects model for health outcomes
feols(
  fphealth ~ head_start | mom_id,
  data = nlsy_kids
)
# Fixed-effects model for high school graduation
feols(
  hsgrad ~ head_start | mom_id,
  data = nlsy_kids
)

# Fixed-effects model for college attendance
feols(
  somecoll ~ head_start | mom_id,
  data = nlsy_kids)

# Fixed-effects model for being idle
feols(
  idle ~ head_start | mom_id,
  data = nlsy_kids)

# Fixed-effects model for having a learning disability
feols(
  learndis ~ head_start | mom_id,
  data = nlsy_kids)

# The results of the mother-fixed effect model show that Head Start participation 
# has a negative, insignificant, close-to-zero effect on whether an individual 
# will repeat a grade in their life. Similarly, the effect on health outcomes
# is very slightly but significantly negative, implying that participation slightly
# decreases the likelihood of reporting fair/poor health in later years. Moreover, 
# the effect of Head Start participation on high school graduation likelihood is small, 
# but significantly positive. Furthermore, the effect of Head Start participation
# on college attendance is small but significantly positive. Second to last, Head Start
# participation has a small but significantly negative effect on the likelihood of 
# being idle. Lastly, Head Start seems to have a small but significantly negative effect
# on the likelihood of being diagnosed with a learning disability.
# Generally, it seems that Head Start participation seems to improve life outcomes 
# slightly but significantly.
```

\newpage

## Question 9

Code: Analysis of heterogeneity.

Verbal: Interpret.

```{r}
# All question 9 code here
# Fixed-effects model with interaction terms for race/ethnicity
feols(
  hsgrad ~ head_start + head_start * black + head_start * hispanic | mom_id,
  data = nlsy_kids)

# Judging from the regression results, it seems that the effect of Head Start
# participation on the likelihood of high school graduation improves for individuals
# that are black and hispanic compared to all other individuals. However, the results
# are insignificant and should thus be treated with caution.

```

\newpage

## Question 10

Verbal: Policy implications.

I would argue that the results support Joe Biden's proposal of expanding federal
funding for early-childhood education programmes. Based on our fixed-effects estimates
we find significantly and large positive effects on test scores at ages 5 to 6, which 
stay significant but shrink in size in later years. Moreover, our estimates for later
life outcomes suggest significant although small positive effects. I would be comfortable
predicting the qualitative effects of such an expansion but not the quantitative
ones based on my results, because the magnitude of the effect does change quite a lot
when excluding individuals with na scores for the covariates in question 6, which implies
that the size of the coefficient seems quite sensitive to the sample that it is being 
estimated on, making me uncomfortable in predicting the quantitive effects of an expansion
based on these results.


